---
title: '[AP] Assignent'
output:
  word_document: default
  html_document: default
date: "2025-03-30"
editor_options:
  markdown:
    wrap: 72
---

# I. Personal Information:

-   Name: Pham Hoang Minh Hien
-   Student ID: 22027464
-   Subject: Annalytic Programming

# II. Declaration:

By including this statement, we the authors of this work, verify that:

-   We hold a copy of this assignment that we can produce if the
    original is lost or damaged.

-   We hereby certify that no part of this assignment/product has been
    copied from any other student’s work or from any other source except
    where due acknowledgement is made in the assignment.

-   No part of this assignment/product has been written/produced for us
    by another person except where such collaboration has been
    authorised by the subject lecturer/tutor concerned.

-   We are aware that this work may be reproduced and submitted to
    plagiarism detection software programs for the purpose of detecting
    possible plagiarism (which may retain a copy on its database for
    future plagiarism checking).

-   We hereby certify that we have read and understand what the School
    of Computing, Engineering and Mathematics defines as minor and
    substantial breaches of misconduct as outlined in the learning guide
    for this unit.

# III. Assignment

## Question 1:

1.  Write the code to inspect the data structure and present the data:

2.  The missing values in the dataset were written as "?", replace any
    "?" with NA;

3.  Convert categorical variables BodyStyles, FuelTypes, ErrorCodes to
    factors;

4.  Replace the missing values in column Horsepower with the mean
    horsepower;

5.  Select the appropriate chart type and display: horsepower
    distribution.

### Loading dataset to R

```{r Data Loading}
auto = read.csv("C:/Users/LENOVO/Downloads/Automobile.csv")
engine = read.csv("C:/Users/LENOVO/Downloads/Engine.csv")
maintenance = read.csv("C:/Users/LENOVO/Downloads/Maintenance.csv")
```

### Inspect dataset

```{r Inspect data}
## Automobile dataset
head(auto)
str(auto)
summary(auto)

## Engine dataset
head(engine)
str(engine)
summary(engine)

## Maintenance  dataset
head(maintenance)
str(maintenance)
summary(maintenance)
```

### Define "?" and replace with NA

```{r}
# Engine dataset
summary(engine == "?")

# Auto dataset
summary(auto == "?")

# Maintenance dataset
summary(maintenance == "?")
```

After scanning through three datasets, there are "?" in Engine dataset.
Therefore, missing values represented by "?" were replaced with NA to
standardize across datasets. .

```{r}
# Replace "?" with NA
engine_new <- engine
engine_new[engine_new == "?"] <- NA

summary(engine_new == "?")
```

### Change factor (Convert categorical variables BodyStyles, FuelTypes, ErrorCodes to factors; )

```{r}
# Inspect the dataset for columns names
names(auto)
names(engine_new)
names(maintenance)

# Change the variables to factors
auto$BodyStyles <- as.factor(auto$BodyStyles)
engine_new$FuelTypes <- as.factor(engine_new$FuelTypes)
maintenance$ErrorCodes <- as.factor(maintenance$ErrorCodes)

# Check the dataset types of variables
str(auto)
str(engine_new)
str(maintenance)
```

### Replace the missing values in column Horsepower with the mean horsepower;

```{r}
# Convert Horsepower to numeric
engine_new$Horsepower <- as.numeric(engine_new$Horsepower)

# Calculate the mean of Horsepower, excluding NA values
hp_mean <- mean(engine_new$Horsepower, na.rm = TRUE)

# Replace NA values in Horsepower with the calculated mean
engine_new$Horsepower[is.na(engine_new$Horsepower)] <- hp_mean

# Check for NA 
summary(is.na(engine_new$Horsepower))
```

After checking the summary of Engine_new data, the missing data in
Horesepower has been replaced to mean Horepower from the whole column.It
creates the continuos variable with a nearly normal distribution for
further analysis.

The EngineType column which has 5 NAs, can not replace, due to the
character variables.

## Select the appropriate chart type and display: horsepower distribution.

```{r}
#Launching the needed library
library("ggplot2")
```

```{r}
ggplot(engine_new, aes(x = Horsepower)) +
  geom_histogram(binwidth = 10, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Horsepower Distribution", x = "Horsepower", y = "Count") +
  theme_minimal()

ggplot(engine_new, aes(x = Horsepower)) +
  geom_histogram(binwidth = 5, fill = "green", color = "black", alpha = 0.7) +
  labs(title = "Horsepower Distribution", x = "Horsepower", y = "Count") +
  theme_minimal()
```

The histograms effectively illustrate how horsepower values are
distributed across different ranges, helping to identify both common
values and outliers. Peaks in the distribution indicate horsepower
ranges with high frequency, while dips highlight less common values.

There is a noticeable difference between the two histograms due to the
different bin widths used: the first histogram uses a binwidth of 10,
while the second uses 5. The first histogram provides a smoother, more
generalized view of the distribution; in contrast, the second histogram
offers a more detailed breakdown.

Overall, the choice of binwidth affects the level of detail shown:
larger bins reveal general trends, while smaller bins highlight finer
variations in the horsepower distribution.

## Question 2

Write the code to analyse the distribution of the horsepower across the
number of cylinders. Write the code to investigate the distribution of
the horsepower across the groups of the engine sizes (e.g., 60-100,
101-200, 201-300, 301+). Visualize both the findings using the
histogram. Explain your findings.

### Call out the Cylinder

```{r}
engine_new$NumCylinders <- as.factor(engine_new$NumCylinders)

#Create visualization
ggplot(engine_new, aes(x = Horsepower, fill = NumCylinders)) +
  geom_histogram(binwidth = 20, color = "black", alpha = 0.6, position = "dodge") +
  labs(title = "Horsepower Distribution Across Cylinders", x = "Horsepower", y = "Count") +
  theme_minimal()
```

**Analysis:**

-   Through the graph "Horsepower Distribution Across Cylinders", the
    most outstanding Cylinder in the dataset is the number of cars which
    have three-cylinder. They appear to be the most common configuration
    which cover a wide range from under 100 to almost 200 HP. In a
    deeper analysed, they primarily concentrated in the range from 75 -
    150 HP, which the highest frequency is almost 100HP.

-   Fourth-Cylinders has the second most popular since they appears
    mostly from 100 - 200 HP ranges. However, the counts of number are
    not out-standingly compared to the Three-cylinder, they have higher
    Horsepower ranges.

-   Other cylinder, such as the number of Five, Six, One have the least
    count due to they have the low range performance (over 50 HP for
    Five-Cylinder) and very high performance that other can not reach
    (approximately 300 HP).

-   Through out the distribution, it shows a clear positive correlation
    between the number of cylinders and horsepower; however, to
    demonstrate the correlation of these two varibales, we can used
    correlation test to identify correctly.

```{r}
# Converting NumCylinders back to numeric for correlation test
engine_new$NumCylinders <- as.numeric(engine_new$NumCylinders)
cor_test <- cor.test(engine_new$NumCylinders, engine_new$Horsepower)
print(cor_test)
```

The p-value of the correlation test is higher than 0.05 (0.7213 \> 0.05)
which indicate that there are not significantly impact each other.

### Group size

Based on the ranges of Engine Size, the dataset provides wide range of
numbers from 60 to over 200. As the example of number cylinders above,
we can divide the Engine Size to 5 groups, with 50 size range each.

```{r}
# Create engine size groups
engine_new$Engine_Size_Group <- cut(engine_new$EngineSize, 
                                breaks = c(60, 100, 150, 200, 300, Inf), 
                                labels = c("60-100","101-159", "151-200", "201-300", "301+"), 
                                include.lowest = TRUE)

# Histogram of Horsepower grouped by Engine Size
ggplot(engine_new, aes(x = Horsepower, fill = Engine_Size_Group)) +
  geom_histogram(binwidth = 20, color = "black", alpha = 0.6, position = "dodge") +
  labs(title = "Horsepower Distribution Across Engine Size Groups", x = "Horsepower", y = "Count") +
  theme_minimal()
```

**Analysis**:

\- From the histogram, there are two groups have high frequency, compare
to other three group that have Engine Size from 150 and above.

\- Comparing the first two groups (Engine Size under 100 and Engine Size
from 101 to 150), Size Under 100 group has a horsepower range from under
50HP to almost 150 HP, which their highest frequency of Horsepower at
50HP with over 15 cars are reported. The group from 101-150 Size has a
wider range, which from 50HP to almost 175HP. They have the highest
reported frequency is from 75HP to 150HP with the peak with over 15 cars
report at 100 HP.

\- Other three engines size groups have largest horpower range, which is
from 100HP and above. With the engine size from 200, the horsepower is
reported to be almost 200HP; Relatively, the car could reach to almost
300HP if the engine are from the rang 300.

\- To test the relationship of the Engine size to the Horsepower
performance, we can use correlation test:

```{r}
# Using original EngineSize column
cor_test2 <- cor.test(engine_new$EngineSize, engine_new$Horsepower)
print(cor_test2)
```

The p-value is small, under 0.05 which can demtermine that Engíne Size
have positve impact to Horsepower performance. Engines have larger size
tend to have larger displacements and produce more power.

## Question 3:

Filter out those engines in the dataset that have trouble or are
suspected of having trouble; What are the top 5 most common troubles
related to the engines? Do the troubles differ between fuel types?
Provide a table to rank the top 5 troubles for diesel and gas engines
separately. Elaborate on the findings.

```{r}
# Load necessary library
library(dplyr)
```

### Create a dataset which contain troubles

```{r}
# call out the maintenance dataset
head(maintenance)

# Filter a new dataset contain engine model and engine type
maintenance_new <- maintenance %>%
  left_join(auto %>% select(PlateNumber, EngineModel), by = "PlateNumber")

engine_new_unique <- engine_new %>%
  arrange(EngineModel) %>%  # Ensure sorted order (optional)
  distinct(EngineModel, .keep_all = TRUE)  # Keep only first occurrence

maintenance_new_2 <- maintenance_new %>%
  left_join(engine_new_unique %>% select(EngineModel, EngineType), by = "EngineModel")

# Remove rows where Troubles is "No Problem"
maintenance_new_2 <- maintenance_new_2 %>%
  filter(Troubles != "No error")

head(maintenance_new_2)
```

After inspecting the engine dataset, there are several duplicate of
engine which have diffences in EngineSize, which can leads to confusion
in Maintenance dataset after joining. Therefore, we create
engine_new_unique which only take the unique engine and remove the
duplicate for better arrangement and analysis.

### Create top 5 most common troubles within the engine

```{r}
# Function to filter out engines with troubles and count trouble occurrences
filter_troubled_engines <- function(df) {
  # Total number of trouble occurrences
  total_troubles <- nrow(df)
  
  # Count occurrences of each trouble and calculate percentage
  top_problems <- df %>% 
    count(Troubles, sort = TRUE) %>% 
    mutate(Percentage = round(n / total_troubles * 100, 2)) %>% 
    head(5)  # Keep top 5
  
  # Count occurrences of each troubled engine and calculate percentage
  top_troubled_engines <- df %>% 
    count(EngineModel, sort = TRUE) %>% 
    mutate(Percentage = round(n / total_troubles * 100, 2)) %>% 
    head(5)  # Keep top 5
  
  # Return results as a list
  list(
    top_5_problems = top_problems,
    top_5_troubled_engines = top_troubled_engines
  )
}

# Apply the function to maintenance_new_2 dataset
trouble_results <- filter_troubled_engines(maintenance_new_2)

# Display results
head(trouble_results$top_5_problems)
head(trouble_results$top_5_troubled_engines)

```

Above the table of top 5 troubles which occurred frequently when car
maintenance. The top trouble was related to "Cylinder", occurrence for
38 times as almost 11%. Following problems were ignition and
noise-related problems.

### Rankings by FuelType

```{r}
# Adding fuel type to the maintenance_new_2 dataset
maintenance_new_3 <- maintenance_new_2 %>% 
  left_join(engine_new_unique %>% select(EngineModel, FuelTypes), by = "EngineModel")


unique(maintenance_new_3$FuelTypes)

maintenance_new_gas <- maintenance_new_3 %>% 
  filter(FuelTypes == "gas")
maintenance_new_diesel <- maintenance_new_3 %>% 
  filter(FuelTypes == "diesel")

# Function to filter out engines with troubles and count trouble occurrences
filter_troubled_by_fueltypes <- function(df) {
  #Filter out the trouble occur within the engine, compare to other vehicle components
  df <- df %>% filter(ErrorCodes == "1")
  # Count occurrences of each trouble
  top_problems <- df %>% count(Troubles, sort = TRUE)
  # Count troubles of each engines
  top_troubled_engines <- df %>% count(EngineModel, sort = TRUE)
  # Return results as a list
  list(
    top_5_problems = top_problems,
    top_5_troubled_engines = top_troubled_engines)
}

## Apply the function
trouble_gas <- filter_troubled_by_fueltypes(maintenance_new_gas)
trouble_diesel <- filter_troubled_by_fueltypes(maintenance_new_diesel)

# Display results
head(trouble_gas$top_5_problems)
head(trouble_diesel$top_5_problems)
```

When split by fuel type, gas engines showed more frequent wear-and-tear
related issues, such as ignition and the most common problem,
“cylinders”. In contrast, diesel engines experienced fewer reported
issues, but these were often mechanically complex, such as cam shafts
and strokes. However, the diesel dataset represents a small proportion
of total vehicles, so direct comparisons may be less reliable.

## Question 4:

Write the code to analyze the factors that might influence the
maintenance methods (Urgent care, Adjustment, Replacement) for the
trouble vehicles (confirmed or suspected) in the dataset. Any factors in
the dataset, such as BodyStyles, FuelTypes, and ErrorCodes, can be
considered. Pick 2 of the factors and explain if there is a trend that
explains the variation.

### Create the dataset accordingly

```{r}
maintenance_q4 <- maintenance_new_3 %>% 
  select(ID, PlateNumber, Troubles, ErrorCodes, Methods, EngineType, FuelTypes) %>% 
  left_join(auto %>% select(PlateNumber, BodyStyles), by = "PlateNumber") %>%
  filter(!is.na(Methods))

str(maintenance_q4)

# Convert variable to factor
maintenance_q4$Methods <- as.factor(maintenance_q4$Methods)

```

### Create summary tables

```{r}
# Summarize categorical variables
table_fuel <- table(maintenance_q4$FuelTypes, maintenance_q4$Methods)
table_body <- table(maintenance_q4$BodyStyles, maintenance_q4$Methods)

print(table_fuel)
print(table_body)
```

### Chi Squared for testing correlation

```{r}
# Chi-square test for FuelTypes and Methods
chi_test_fuel <- chisq.test(table_fuel)
print(chi_test_fuel)

# Chi-square test for BodyStyles and Methods
chi_test_body <- chisq.test(table_body)
print(chi_test_body)

```

The distribution showed that Fuel Type has a statistically significant
relationship with the type of maintenance method (p = 0.047), where gas
vehicles more frequently received “Urgent care” than diesel. Conversely,
BodyStyle had no significant influence on the method used (p = 0.65).

### Visualization - Distribution of Methods by Fuel Type

```{r}
# Plot by Fuel Type
ggplot(maintenance_q4, aes(x = FuelTypes, fill = Methods)) +
  geom_bar(position = "dodge") +
  labs(title = "Maintenance Methods by Fuel Type", x = "Fuel Type", y = "Count") +
  theme_minimal()
```

Across all body styles, “Replacement” is the most common maintenance
method, indicating it's the go-to solution regardless of car type.

```{r}
# Plot by Body Style
ggplot(maintenance_q4, aes(x = BodyStyles, fill = Methods)) +
  geom_bar(position = "dodge") +
  labs(title = "Maintenance Methods by Body Style", x = "Body Style", y = "Count") +
  theme_minimal()
```

Sedans and hatchbacks account for the majority of maintenance cases and
show a wider spread across methods, including a higher count of “Urgent
care”. Convertibles and hardtops have very few maintenance cases
overall, with almost no urgent care observed. This may suggest that
common, high-usage vehicles (like sedans and hatchbacks) experience more
varied and possibly severe issues, while less common body styles may
require simpler, less frequent maintenance.
